---
layout: post
title: Data anonymization in R
tags: R digest data.table
---

# Use cases

* Public reports.  
* Public data sharing, e.g. R packages download logs from CRAN's RStudio mirror - [cran-logs.rstudio.com](http://cran-logs.rstudio.com/) - mask ip addresses.
* Reports or data sharing for external vendor.
* Development works can operate on anonymized PRODUCTION data.  
Manually or semi-manually populated data can often brings some new issue after migration to PRODUCTION data.  
Such anonymized PRODUCTION data can be quite handy for the devs.  

# Dependencies

```{r init, echo=c(1)}
suppressPackageStartupMessages({
  library(data.table)
  library(digest)
  library(knitr) # used only for post creation
})
# # produce md
# knitr::knit("2014-11-07-Data-Anonymization-in-R.Rmd")
opts_chunk$set(collapse=TRUE, comment="#>", cache=FALSE)
kable <- function(x, ..., digits=2, longtable=TRUE, col.names=gsub("\\b([a-z])", "\\U\\1",gsub("_"," ",colnames(x)), perl=TRUE)) knitr::kable(x=x, ..., digits=digits, longtable=longtable, col.names=col.names)
set.seed(1)
```

# Sample of survey data

Anonymize sensitive information in survey data, data storage in a single table.
```{r data_SURV, echo=FALSE}
SURV <- list()
SURV[[length(SURV)+1]] <- data.table(city = "London",postal_code = "SW1H 0QW",address = "Silk Road 17",
                                     name = "John Lennon",sex = "M",
                                     age = 48L,height = 176,weight = 94,score = 3L)
SURV[[length(SURV)+1]] <- data.table(city = "Cardiff",postal_code = "CF23 9AE",address = "Queen Road 19",
                                     name = "Edward Snowden",sex = "M",
                                     age = 55L,height = 185,weight = 74,score = 2L)
SURV[[length(SURV)+1]] <- data.table(city = "London",postal_code = "SW1P 3BU",address = "Edinburgh Road 19",
                                     name = "John Kennedy",sex = "M",
                                     age = 46L,height = 156,weight = 84,score = 1L)
SURV[[length(SURV)+1]] <- data.table(city = "London",postal_code = "SW1P 3BU",address = "Cardiff Road 21",
                                     name = "Mahatma Gandhi",sex = "M",
                                     age = 56L,height = 186,weight = 54,score = 5L)
SURV[[length(SURV)+1]] <- data.table(city = "Cardiff",postal_code = "CF23 9AE",address = "King Road 10",
                                     name = "Nelson Mandela",sex = "M",
                                     age = 61L,height = 181,weight = 84,score = 2L)
SURV[[length(SURV)+1]] <- data.table(city = "London",postal_code = "SW1P 2EE",address = "Cardiff Road 23",
                                     name = "Vandana Shiva",sex = "F",
                                     age = 41L,height = 192,weight = 64,score = 5L)
SURV[[length(SURV)+1]] <- data.table(city = "London",postal_code = "SW1P 2EE",address = "Cardiff Road 23",
                                     name = "Yasser Arafat",sex = "M",
                                     age = 32L,height = 190,weight = 85,score = 2L)
SURV[[length(SURV)+1]] <- data.table(city = "Cardiff",postal_code = "CF23 9AE",address = "Queen Road 9",
                                     name = "Rachel Corrie",sex = "F",
                                     age = 80L,height = 155,weight = 77,score = 3L)
SURV[[length(SURV)+1]] <- data.table(city = "Glasgow",postal_code = "G40 3AS",address = "Simple Road 11",
                                     name = "Bob Marley",sex = "M",
                                     age = 53L,height = 180,weight = 71,score = 2L)
SURV <- rbindlist(SURV)
```
```{r preview_SURV}
# pretty print
kable(head(SURV))
```

# Anonymize function

Function will calculate hashes only for unique inputs and return vector of masked inputs.  
My version will use `digest(x, algo="crc32")` because it fits better into html tables, algo `crc32` is not really secure.  
Read `?digest::digest` for supported `algo`, also consider to salt your input vector, e.g. `x=paste0("prefix",x,"suffix")`.  
Performance improvement possible using `Rcpp` / `C`: [digest #2](https://github.com/eddelbuettel/digest/issues/2).
```{r anonymize_FUN, echo=TRUE, eval=TRUE}
anonymize <- function(x, algo="crc32"){
  unq_hashes <- vapply(unique(x), function(object) digest(object, algo=algo), FUN.VALUE="", USE.NAMES=TRUE)
  unname(unq_hashes[x])
}
```

# Anonymize survey data

We will keep *city* and *sex* fields unmasked.
```{r anonymize_SURV, echo=TRUE}
# choose columns to mask
cols_to_mask <- c("name","address","postal_code")
# backup original data
SURV_ORG <- copy(SURV)
# anonymize
SURV[,cols_to_mask := lapply(.SD, anonymize),.SDcols=cols_to_mask,with=FALSE]
# pretty print
kable(head(SURV))
```

# Why not just random data or integer sequence

When using the `digest` function to hide sensitive data you:

* keep rows distribution:  
aggregates by masked columns will still match to aggregates on original columns, see simple grouping below:
```{r anonymize_SURV_grouping}
SURV_ORG[,.(.N,mean_age=mean(age),mean_score=mean(score)),by=.(city,postal_code)
         ][,kable(.SD)]
SURV[,.(.N,mean_age=mean(age),mean_score=mean(score)),by=.(city,postal_code)
     ][,kable(.SD)]
```
* keep relationships on equi joins:  
if `t1.col1 == t2.col4` TRUE then also `digest(t1.col1) == digest(t2.col4)` TRUE  
Example in next section below.

# Sample of sales data

Anonymize relational data in sales data, data normalized into *SALES* and *CUSTOMER* tables.
```{r SALES, echo=FALSE}
CUSTOMER <- copy(SURV_ORG[,.(customer_uid = paste("CUST",.I,sep="_"),city,postal_code,address,name,sex)])
setkey(CUSTOMER,customer_uid)
N <- 100
SALES <- data.table(customer_uid=sample(CUSTOMER$customer_uid, size=N, replace=TRUE), product_name=paste0(sample(letters,N,TRUE),sample(letters,N,TRUE),sample(letters,N,TRUE)), transaction_date=sample(seq(Sys.Date()-30,Sys.Date(),by=1),N,TRUE), quantity=quantity<-sample(1:60,50,TRUE), value=quantity*sample(5:20,50,TRUE))
```

```{r preview_SALES}
kable(head(SALES,4))
kable(head(CUSTOMER,2))
```
```{r SALES_join}
# join
kable(head(
  CUSTOMER[SALES]
))
```
```{r SALES_join_aggr}
# join and aggregate
kable(head(
  CUSTOMER[SALES][,.(quantity = sum(quantity),value = sum(value)),by=.(city,postal_code)]
))
```

# Anonymize sales data

```{r anonymize_SALES}
SALES[, customer_uid := anonymize(customer_uid)]
cols_to_mask <- c("customer_uid","name","address","postal_code")
CUSTOMER[,cols_to_mask := lapply(.SD, anonymize),.SDcols=cols_to_mask,with=FALSE]
setkey(CUSTOMER,customer_uid)
```

```{r preview_anonymize_SALES}
# preview result
kable(head(CUSTOMER,2))
kable(head(SALES,2))
```

```{r anonymize_SALES_join}
# datasets will still join correctly even on masked columns
kable(head(
  CUSTOMER[SALES]
))
```

```{r anonymize_SALES_join_aggr}
# also the aggregates on masked columns will match to the origin
kable(head(
    CUSTOMER[SALES][,.(quantity = sum(quantity),value = sum(value)),by=.(city,postal_code)]
))
```

# Reproduce from Rmd

Script used to produce this post is available in the github repo (link in the page footer) as `Rmd` file and can be easily reproduced locally in R (required [knitr](http://cran.r-project.org/web/packages/knitr/index.html) or [rmarkdown](http://cran.r-project.org/web/packages/rmarkdown/index.html)) to any format (`md`, `html`, `pdf`, `docx`).
```{r reproduce_Rmd, eval=FALSE}
# html output
rmarkdown::render("2014-11-07-Data-Anonymization-in-R.Rmd", html_document())
# markdown file used as current post
knitr::knit("2014-11-07-Data-Anonymization-in-R.Rmd")
```

# Minimal script

Minimal script example on survey data as `SURV_ORG` data.table:

```{r min_script, eval=FALSE}
anonymize <- function(x, algo="crc32"){
  unq_hashes <- vapply(unique(x), function(object) digest(object, algo=algo), FUN.VALUE="", USE.NAMES=TRUE)
  unname(unq_hashes[x])
}
cols_to_mask <- c("name","address","postal_code")
SURV_ORG[, cols_to_mask := lapply(.SD, anonymize), .SDcols=cols_to_mask, with=FALSE][]
```
