---
layout: post
title: Auditing data transformation
tags: R ETL data.table
---

```{r init, echo=FALSE}
# knitr::knit("_scripts/2015-06-03-dtq.Rmd", "_posts/2015-06-03-dtq.md")
options(width=100)
```

Auditing data transformation can be simply described as gathering metadata about the transformation process. The most basics metadata would be a timestamp, atomic transformation description, data volume on input, data volume on output, time elapsed.  

If you work with R only interactively you may find it more like a fancy tool. On the other hand for automated scheduled R jobs it may be quite helpful to have traceability on the lower grain of processing than just binary *success* or *fail* after the script is executed, for example a logging each query against the data.  
Similar features are already available in [ETL](https://en.wikipedia.org/wiki/Extract,_transform,_load) tools for decades.  

### Implementation

I've addressed this in my [dtq](https://github.com/jangorecki/dtq) package.  
It silently eavesdrop every call to `[` for any `data.table`.  
It can be perceived as preconfigured and faster version of `base::trace` tailored for data transformation auditing.  

I believe it should be possible to port the idea to [magrittr](https://github.com/smbache/magrittr)'s `%>%` call if somebody need auditing there.  

### Examples

Populate example sales data.  

```{r data, message=FALSE}
library(dtq)
set.seed(1)
DT <- data.table(
  user = 1:10,
  group = letters[1:5],
  time = as.POSIXct(seq(from=1.4e9L+1L,to=1.5e9L,by=1e3), origin="1970-01-01"),
  sales = rnorm(10*5*1e5, 5)
)
knitr::kable(head(DT, 2))
```

#### Basic example

Order data, various aggregations.  

```{r examples_basic, results='hide'}
## just to clarify below usage of unnamed args in `[`
# DT[ i, j, by, keyby ]

# first sale by user in each month
DT[order(time), head(.SD, 1), .(user, year(time), month(time))]
# total sales by user in each month
DT[, .(sales = sum(sales)), .(user, year(time), month(time))]
# total sales by group of users in each months, returning ordered and keyed data
DT[, .(sales = sum(sales)),, .(group, year(time), month(time))]
```

Lets preview transformation metadata recorded by `dtq`.  

```{r log_basic}
knitr::kable(dtl(print = TRUE))
```

#### Complex example

Populate budget data to set together with sales data.  

```{r examples_complex_data, results='hide'}
BUDGET <- CJ(
  group = letters[1:5],
  year = 2014:2016,
  month = 1:12
) # cross of dimensions
BUDGET[, budget := rnorm(3*12*5, 12*1e4, 12*1e3)] # budget value measure
```
  
  
> For more complex processing I've found it very useful to organize single conceptual query on data into single chain of `[` calls.  

In below example I will calculate *year-to-date* sales and budget values and *last year* sales vs current sales ratio.  

```{r examples_complex, results='hide'}
# column names
measures <- c("sales","budget")
ytd_measures <- paste("ytd", measures, sep="_")
ly_ytd_measures <- paste("ly", ytd_measures, sep="_")
# query
DT[, .(sales = sum(sales)),, .(group, year = year(time), month = month(time))
   ][BUDGET, `:=`(budget = i.budget, budget_ratio = sales / i.budget)
     ][, c(ytd_measures) := lapply(.SD, cumsum), .(group, year), .SDcol = measures
       ][, c(ly_ytd_measures) := shift(.SD), .(group, month), .SDcol = ytd_measures
         ][, ytd_sales_vs_ly_ytd_sales := ytd_sales / ly_ytd_sales
           ]
```

Check `dtq` logs again.  

```{r log_complex}
knitr::kable(dtl(print = TRUE))
```

### Minimal overhead

Time added to transformation due to logging using `dtq` is minimal.  
For time consuming queries it is unnoticeable ~ 0% overhead.  
For the most pessimistic, unrealistic query: `for(i in seq_len(1e3)) data.table(a=1L)[,.(a)]` it adds around 30% overhead.  

Processes evaluated by `dtq` on each data.table `[` call:  

- gathering information
    - current *call*
    - top *environment* of the current call
    - `nrow` on input and output
    - `proc.time` on start and end, or `get_nanotime` if available
- append gathered info to log storage
    - `c` function is used to append current log to existing logs

All other information (*sources*, *sequences*, *queries*) are extracted from `dtq` logs while accessing them by `dtl()`. So they are not adding any overhead during data transformation.  
The `dtl()` is a function to query against [R6 class](https://github.com/wch/R6) log storage.  

### Using in package

It is common to form your scripts into package so you may want to track data transformation from within your package.  
It is handled by default, you don't need anything more than `library(dtq)` on script init.  
Actually you can restrict logging to particular package or global env, which is useful if some of your dependencies are also using data.table and by default would be logged together with your logs.  

### Example use cases

#### Storing dtq logs in rds/csv/db

Traceability is good but if you close R session the `dtq` logs will be gone.  
Simplest would be `write.table` with `append=TRUE` on the end of script.  
Use `dtl(print = TRUE)` to get rid off list-type columns which stores the `R6` or `call` objects.  
If you want to keep R data types you should then use `saveRDS` on `dtl()`.  

#### Performance tuning  

You can easily analyse your logs and detect the most time consuming sub-queries. Compare its timing to volume of in/our rows, etc.  

#### Apply business rules on metadata

Your currency exchange source system is commited to provide daily exchanges rates in 5 minutes intervals.  

 - You expect exactly 288 rows on the input. If that is not true you catch it and send email directly from R to source system support.  
 - In your high level processing you expect to have 5 different KPIs on output. If it is not true you catch it and send email to yourself.  

### PS.

I would appreciate if someone could recommend a commenting service for github jekyll blogs, without tracking *feature* like in disqus service. Including self hosted of course.  
